---
title: "Course 5 Programming Assignment 1"
output:
  md_document:
    variant: markdown_github
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Reading in the Data

```{r reading}
dfActivity<-read.csv("activity.csv")
str(dfActivity)
```

## Mean Total Number of Steps Taken Per Day

Histogram & Mean Calculations <br />
Ignoring NAs <br />

```{r mean}

totalSteps<-tapply(X = dfActivity$steps,INDEX = dfActivity$date,sum,na.rm=TRUE)
hist(x = totalSteps,main = "Frequency of Total Steps Taken ", xlab = "Total Steps Per Day")
print(paste0("Mean: ",mean(totalSteps,na.rm = T)))
print(paste0("Median: ",median(totalSteps,na.rm = T)))
```

## Average Daily Activity Pattern

Time Series Plot <br />
Interval with the heighest  <br />

```{r average}

averageSteps<-tapply(X = dfActivity$steps,INDEX = dfActivity$interval, mean, na.rm=TRUE)
plot(averageSteps,type="l",col="purple",xlab="Interval",ylab="Average Number of Steps")
print(paste0("Interal with the highest average number of steps: ", averageSteps[which.max(averageSteps)]))

```

## Dealing with Missing Values

```{r missing}

print(paste0("Number of rows with NA's: ", sum(!complete.cases(dfActivity))))

dfActivity2<-dfActivity

## Fill in missing data with averages of the matching interval
for(i in 1:nrow(dfActivity2)){
    if(is.na(dfActivity2$steps[i])){
        for(j in 1:nrow(averageSteps)){
            if(as.character(dfActivity2$interval[i])==names(averageSteps[j])){
                dfActivity2$steps[i] = averageSteps[[j]]
            }
        }
    }
}

print(paste0("Number of rows left with NA's: ", sum(!complete.cases(dfActivity2))))

```

```{r mean2}

totalSteps2<-tapply(X = dfActivity2$steps,INDEX = dfActivity2$date,sum)
hist(x = totalSteps2, xlab = "Total Steps Per Day", main="Total Number of Steps with Missing Values Addressed")

print(paste0("Mean of original DF: ",mean(totalSteps,na.rm = T)))
print(paste0("Median of original DF: ",median(totalSteps,na.rm = T)))

print(paste0("Mean of complete DF: ",mean(totalSteps2)))
print(paste0("Median of complete DF: ",median(totalSteps2)))
```

##### By estimating the missing values, both the mean and median increased. It also caused the histogram to boost up the middle bucket compared to the other buckets. 


## Differences Between Weekdays and Weekends

```{r weekend}

dfActivity$days <- weekdays(as.POSIXlt.date(dfActivity$date))
dfActivity$daytype <- dfActivity$days

for(k in 1:nrow(dfActivity)){
    if(dfActivity$days[k] == "Saturday" || dfActivity$days[k] == "Sunday"){
        dfActivity$daytype[k] = "weekend"
    } else {
        dfActivity$daytype[k] = "weekday"
    }
}

dfActivity$daytype<-as.factor(dfActivity$daytype)


dfActivityWeekend<-subset(dfActivity,dfActivity$daytype == "weekend")
dfActivityWeekday<-subset(dfActivity,dfActivity$daytype == "weekday")

averageStepsWeekend<-tapply(X = dfActivityWeekend$steps,INDEX = dfActivityWeekend$interval, mean, na.rm=TRUE)
averageStepsWeekday<-tapply(X = dfActivityWeekday$steps,INDEX = dfActivityWeekday$interval, mean, na.rm=TRUE)

par(mfrow = c(2,1))
plot(averageStepsWeekday,type="l",col="red",xlab="Interval",ylab="Steps",main="Average Number of Steps for Weekdays")
plot(averageStepsWeekend,type="l",col="blue",xlab="Interval",ylab="Steps",main="Average Number of Steps for Weekends")

```